Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-03-09 Ben_Manuscripts/hdphmm/Outline.tex
Comments.

  There is a need for highly selective membranes in order to perform efficient 
  separations of components of complex aqueous streams.
  \begin{itemize}
    \item Organic micropollutants
    \item Desalination and boric acid removal from seawater.
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

@@

  Lyotropic liquid crystals (LLC) are a class of amphiphilic molecules whose ordered phases
  can be cross-linked into mechanically strong membranes capable of highly selective
  separations.
  \begin{itemize}
    \item The shape of the LLC monomers and water content dictates the ordered phase 
    that they form. There are two phases of particular interest for membrane applications.
  	\item H\textsubscript{II} phase lyotropic liquid crystals are characterized by 
  	hexagonally packed, straight, pores while the Q\textsubscript{I} phase consists of
  	a tortuous network of 3D interconnected pores. 
  	\item In both cases the pores are uniform in size with radii on the order of 1 nm
  	giving them a very strict molecular weight cut-off.
	\item Additionally, they have the potential to disrupt conventional membrane separation
	techniques by being selective based not only on size and charge, but on chemical
	functionality as well.
	\item Their pores are lined with LLC monomer functional groups which can potentially
	be designed to interact with solutes in a chemically-specific manner
  \end{itemize}

  There are limits to what we can learn from experiment about LLC membrane design.
  \begin{itemize}
    \item Experimental observables like permeability and selectivity allow us to speculate
    about the molecular origins of separation processes.
    \item This drives an empirical design approach which can potentially neglect key
    interactions which influence selectivity.
    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
    explained by relatively simple macroscopic models. 
  \end{itemize}

@@

  Unfortunately, the timescales that we can simulate with MD are insufficient to be
  able to make well-converged predictions of macroscopic transport properties 
  traditionally used to characterize membranes in the lab.
  \begin{itemize}
    \item However, if we use descriptive stochastic models that can capture solute
    dynamics, then we could project long timescale behavior in addition to gaining
    a deeper understanding of solute behavior on short timescales.
  \end{itemize}

@@

  % BJC: Not sure whether to call it the infinite hidden markov model or hierarchical dirichlet process hidden markov model
  % The former is definitely simpler.
  % MRS: True.  I would probably see which is used more in the literature. 
  % BJC1: going with IHMM. HDPHMM seems to mostly be used by Emily Fox
  In this work, we apply the infinite hidden markov model (IHMM), a modeling
  approach that is agnostic to the source of time series data, in order to 
  automatically detect and infer the parameters of an unknown number of latent
  autoregressive (AR) modes present in solute center-of-mass time series.
  \begin{itemize}
  	\item In addition to AR parameters for each state, the IHMM estimates the
  	state transition probability matrix.
  	\item The model helps simultaneously uncover underlying transport mechanisms which
  	give rise to dynamical behavior and project that behavior on longer timescales so
  	that we can estimate macroscopic transport observables.
  \end{itemize}

  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms.
  \begin{itemize}
    \item We compare the inferred mechanisms to those which we manually
    identified in our previous work. 
    \item Some kind of conclusion here. Did we find more or less states. Any new states/ subdivisions of states?
  \end{itemize}

@@

  Finally, we use the stochastic trajectory realizations in order to compute the 
  macroscopic flux of each solute and selectivity of the LLC membranes studied towards
  each solute.
  \begin{itemize}
    \item We relate these macroscopic properties to our nanoscopic model by simulating 
    mean first passage time (MFPT).
    \item Some kind of conclusion. This membrane is selective towards solutes with this
    functionality. 
    \item Does the conclusion agree with our previous work? Any length dependence? (I think not)
  \end{itemize}

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10 \% water by weight. 
  \begin{itemize}
    \item Approximately one third of the water molecules occupy the tail region 
    with the rest near the pore center.
    \item We chose to study the 10 wt \% water system because solutes move 
    significantly faster than in the 5 wt \% system studied previously.
    \item Appropriate stochastic modeling requires that solutes sample the 
    accessible mechanisms with representative probability.
  \end{itemize}

@@

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends only on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item Unfortunately, standard HMMs require the number of hidden states to be known
    a priori.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

@@

  %BJC: I do need to include this
  \noindent We describe the dynamics of each state using a vector autoregressive (VAR) model. 
  \begin{itemize}
  	\item A VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim N(0, \Sigma)
  	\end{equation}
  	Previous observations are weighted by coefficient matrices $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  	\item We assumed multivariate Gaussian noise, with mean zero and covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$. %BJC: might try higher orders, but probably not worth it
  	\item We used a conjugate matrix-normal inverse-Wishart prior on parameters
  	$A$ and $\Sigma$ and a conjugate Gaussian prior on $\mathbf{c}$ in order to analytically
  	draw from the posterior.~\cite{fox_nonparametric_2009}
  \end{itemize}

@@

  We applied the IHMM algorithm to the radial and axial coordinates of each 
  solute center-of-mass trajectory.
  \begin{itemize}
    \item We measured the radial coordinate as the distance from the closest
    pore center.
    \item Because the pores are somewhat tortuous, we approximated the pore 
    center as a spline. See previous work.
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
    \item Since the mean vector, $\mathbf{c}$ of each state can take on 
    continuous values, the IHMM algorithm tends to find a large number of states,
    with very few states revisited.
    % BJC1: This will need to be updated depending on how I cluster. Might need to describe
    % multiple clustering approaches.
    \item However, $\mathbf{A_i}$ and $\mathbf{\Sigma}$ tend to share similarities with
    independently identified states. Therefore, we clustered based on $\mathbf{A_i}$
    and $\mathbf{\Sigma}$, ignoring the mean.
    \item Since  we do not know the number of similar states beforehand, we used
    a non-parametric Bayesian Gaussian mixture model % cite scikit-learn
    \item We used the clustered parameters in order to relabel states in the estimated
    state sequences and to recalculate the state transition probability matrix.
  \end{itemize}

@@

  \section{Results and Discussion}

  \subsection{Inferring Solute Transport Mechanisms}

@@

  \noindent Trajectory realizations qualitatively match MD simulation trajectories.
  \begin{itemize}
    \item Look for hopping and trapping behavior
  \end{itemize}

  \noindent MSDs generated from stochastic trajectories match those from MD.
  \begin{itemize}
  	\item Look at curvature and 1-$\sigma$ confidence intervals
  \end{itemize}

@@

  \noindent The IHMM is not limited to the H\textsubscript{II} phase.

@@

  Molecular simulations were performed using the Extreme Science and
  Engineering Discovery Environment (XSEDE), which is supported by National
  Science Foundation grant number ACI-1548562. Specifically, it used the Bridges
  system, which is supported by NSF award number ACI-1445606, at the Pittsburgh
  Supercomputing Center (PSC). This work also utilized the RMACC Summit supercomputer,
  which is supported by the National Science Foundation (awards ACI-1532235 and
  ACI-1532236), the University of Colorado Boulder, and Colorado State
  University. The Summit supercomputer is a joint effort of the University of
  Colorado Boulder and Colorado State University.

>>>>

  There is a need for highly selective membranes in order to perform efficient 
  separations of components of complex aqueous streams.
  %MRS2: add some more examples (When you get around to it).
  \begin{itemize}
    \item Organic micropollutants
    \item Desalination and boric acid removal from seawater.
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

@@

  Lyotropic liquid crystals (LLC) are a class of amphiphilic molecules whose ordered phases
  can be cross-linked into mechanically strong membranes capable of highly selective
  separations.
  \begin{itemize}
    \item The shape of the LLC monomers and water content dictates the ordered phase 
    that they form. There are two phases of particular interest for membrane applications.
  	\item H\textsubscript{II} phase lyotropic liquid crystals are characterized by 
  	hexagonally packed, straight, pores while the Q\textsubscript{I} phase consists of
  	a tortuous network of 3D interconnected pores. 
  	\item In both cases the pores are uniform in size with radii on the order of 1 nm
         %MRS2: is it really a molecular weight cutoff?
  	giving them a very strict molecular weight cut-off.
	\item Additionally, they have the potential to disrupt conventional membrane separation
	techniques by being selective based not only on size and charge, but on chemical
	functionality as well.
	\item Their pores are lined with LLC monomer functional groups which can potentially
	be designed to interact with solutes in a chemically-specific manner
  \end{itemize}

  There are limits to what we can learn from experiment about LLC membrane design.
  \begin{itemize}
    \item Experimental observables like permeability and selectivity allow us to speculate
    about the molecular origins of separation processes.
    %MRS2: not quite sure what you mean below?
    \item This drives an empirical design approach which can potentially neglect key
    interactions which influence selectivity.
    \item LLC membranes have been shown to exhibit selectivities which cannot be fully 
    explained by relatively simple macroscopic models. 
  \end{itemize}

@@

  Unfortunately, the timescales that we can simulate with MD are insufficient to be
  able to make well-converged predictions of macroscopic transport properties 
  traditionally used to characterize membranes in the lab.
  \begin{itemize}
    %MRS2: would be good to get this next point into the first line of a paragraph, since it's quite important. 
    \item However, if we use descriptive stochastic models that can capture solute
    dynamics, then we could project long timescale behavior in addition to gaining
    a deeper understanding of solute behavior on short timescales.
  \end{itemize}

@@

  % BJC: Not sure whether to call it the infinite hidden markov model or hierarchical dirichlet process hidden markov model
  % The former is definitely simpler.
  % MRS: True.  I would probably see which is used more in the literature. 
  % BJC1: going with IHMM. HDPHMM seems to mostly be used by Emily Fox
  In this work, we apply the infinite hidden Markov Model (IHMM), a modeling
  approach that is agnostic to the source of time series data, in order to 
  automatically detect and infer the parameters of an unknown number of latent
  autoregressive (AR) modes present in solute center-of-mass time series.
  \begin{itemize}
  	\item In addition to AR parameters for each state, the IHMM estimates the
  	state transition probability matrix.
  	\item The model helps simultaneously uncover underlying transport mechanisms which
  	give rise to dynamical behavior and project that behavior on longer timescales so
  	that we can estimate macroscopic transport observables.
  \end{itemize}

  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms.
  \begin{itemize}
    \item We compare the inferred mechanisms to those which we manually
    identified in our previous work. 
    %MRS2: don't put the conclusions in the introduction.  Instead, state clearly the questions you will ask and the hypotheses you will test.
   \item Some kind of conclusion here. Did we find more or less states. Any new states/ subdivisions of states?
  \end{itemize}

@@

  Finally, we use the stochastic trajectory realizations in order to compute the 
  macroscopic flux of each solute and selectivity of the LLC membranes studied towards
  each solute.
  \begin{itemize}
    %MRS2: shouldn't reach conclusions in intro, just set up questions to be answered (conclusions SHOULD be in abstract).
    \item We relate these macroscopic properties to our nanoscopic model by simulating 
    mean first passage time (MFPT).
    \item Some kind of conclusion. This membrane is selective towards solutes with this
    functionality. 
    \item Does the conclusion agree with our previous work? Any length dependence? (I think not)
  \end{itemize}

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10 \% water by weight. 
  \begin{itemize}
    \item Approximately one third of the water molecules occupy the tail region 
    with the rest near the pore center.
    \item We chose to study the 10 wt \% water system because solutes move 
    significantly faster than in the 5 wt \% system studied previously.
    \item Appropriate stochastic modeling requires that solutes sample the 
    accessible mechanisms with representative probability.%MRS: only possible on the ~1-10 \mu scale with relatively fast solutes.
  \end{itemize}

@@

  Hidden Markov models (HMMs) are a useful and widely used technique
  for modeling sequences of observations where the probability of the next observation
  in a sequence depends only on a previous unobserved, latent or hidden, state.~\cite{beal_infinite_2002}
  %MRS2: it can depend on other things, but ALSO depends on the latent state.
  \begin{itemize}
    \item In the context of our simulations, the observations correspond to 
    the center of mass coordinates of the solutes versus time, and the states
    correspond to the dynamical behavior which give rise to those types
    of observations.
    \item Unfortunately, standard HMMs require the number of hidden states to be known
    a priori.
    \item One can partially overcome this by testing a range of numbers of 
    hidden states and determining which is the best representation of the
    data.
  \end{itemize}

@@

  %BJC: I do need to include this
  \noindent We describe the dynamics of each state using a vector autoregressive (VAR) model. 
  \begin{itemize}
  	\item A VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim N(0, \Sigma)
  	\end{equation}
  	Previous observations are weighted by coefficient matrices $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
  	\item We assumed multivariate Gaussian noise, with mean zero and covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$. %BJC: might try higher orders, but probably not worth it
  	\item We used a conjugate matrix-normal inverse-Wishart prior on parameters
  	$A$ and $\Sigma$ and a conjugate Gaussian prior on $\mathbf{c}$ in order to analytically
  	draw from the posterior.~\cite{fox_nonparametric_2009}
  \end{itemize}

@@

  We applied the IHMM algorithm to the radial and axial coordinates of each 
  solute center-of-mass trajectory.
  \begin{itemize}
    \item We measured the radial coordinate as the distance from the closest
    pore center.
    \item Because the pores are somewhat tortuous, we approximated the pore 
    center as a spline. See previous work.
    %MRS2: approximated the pore center line with a spline?
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
    %MRS2: can you enforce that the mean in the z-component is zero?
    %MRS2: I'm trying to think what other constraints should be enforced. 
    %MRS2: the issue with state clustering is that when transitionin between two states that are clustered, 
    %MRS2: it destroys the correlation that would have been within if it was all the same statee.
    %MRS2: I'm not sure what the implications are. If the simulations are RUN with grouped states, 
    %MRS2: then the correlation can be reintroduced, so maybe it doesn't matter?
   \item Since the mean vector, $\mathbf{c}$ of each state can take on 
    continuous values, the IHMM algorithm tends to find a large number of states,
    with very few states revisited.
    % BJC1: This will need to be updated depending on how I cluster. Might need to describe
    % multiple clustering approaches.
    \item However, $\mathbf{A_i}$ and $\mathbf{\Sigma}$ tend to share similarities with
    independently identified states. Therefore, we clustered based on $\mathbf{A_i}$
    and $\mathbf{\Sigma}$, ignoring the mean.
    \item Since  we do not know the number of similar states beforehand, we used
    a non-parametric Bayesian Gaussian mixture model % cite scikit-learn
    \item We used the clustered parameters in order to relabel states in the estimated
    state sequences and to recalculate the state transition probability matrix.
  \end{itemize}

@@

  \section{Results and Discussion}
  %MRS2: how stable are the results to realizations of the system
  \subsection{Inferring Solute Transport Mechanisms}

@@

  \noindent Trajectory realizations qualitatively match MD simulation trajectories.

  \begin{itemize}
    %MRS2: think about how to quantify the simiilarity of the hopping and trapping behavior.
    \item Look for hopping and trapping behavior
  \end{itemize}

  \noindent MSDs generated from stochastic trajectories match those from MD.
  \begin{itemize}
    %MRS2: again, think about the best ways to quantify similar between MD and stochastic trajectories.
  	\item Look at curvature and 1-$\sigma$ confidence intervals
  \end{itemize}

@@

  %MRS2: below probably doesn't need to be stated?
  \noindent The IHMM is not limited to the H\textsubscript{II} phase.

@@

  %MRS2: add GAANN and PRF here to cover bases.
  Molecular simulations were performed using the Extreme Science and
  Engineering Discovery Environment (XSEDE), which is supported by National
  Science Foundation grant number ACI-1548562. Specifically, it used the Bridges
  system, which is supported by NSF award number ACI-1445606, at the Pittsburgh
  Supercomputing Center (PSC). This work also utilized the RMACC Summit supercomputer,
  which is supported by the National Science Foundation (awards ACI-1532235 and
  ACI-1532236), the University of Colorado Boulder, and Colorado State
  University. The Summit supercomputer is a joint effort of the University of
  Colorado Boulder and Colorado State University.
